---
title: "EDS241: Assignment 1"
author: "Marie Rivers"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: 
  pdf_document:
    toc: false
    number_sections: yes
header-includes:
  - \setlength{\parindent}{1em}
  - \usepackage{float}
--- 
  
``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}

# set default chunk options
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, 
                      echo = TRUE, message = FALSE, warning = FALSE)

# load packages
packages=c("stargazer", "here", "tidyr", "dplyr","stringr", "janitor", 
           "cowplot", "ggplot2", "tinytex", "datasets", "tibble", "tidyverse", "readxl", "estimatr", "car")

for (i in packages) {
  if (require(i,character.only=TRUE)==FALSE) {
    install.packages(i,repos='http://cran.us.r-project.org')
  }
  else {
    require(i,character.only=TRUE)
  }
}

#devtools::install_github('rstudio/rmarkdown')
options(scipen=10) # not scientific notation
```

\noindent In this assignment, we use air quality data in R to investigate the relationship between PM2.5 and low birth weight in California. The data came from CalEnviroScreen 4.0, a mapping and data tool produced by the California Office of Environmental Health Hazards Assessment (OEHHA). The data are compiled and constructed from a variety of sources and cover all 8,035 census tracts in California. Source: [https://oehha.ca.gov/calenviroscreen/report/calenviroscreen-40](https://oehha.ca.gov/calenviroscreen/report/calenviroscreen-40)

# Read and Clean Data
```{r warning=FALSE}
data <- read_xlsx(here("data", "CES4.xlsx"), sheet = "CES4.0FINAL_results") %>% 
  rename(CensusTract = "Census Tract", TotalPopulation = "Total Population", CaliforniaCounty = "California County", LowBirthWeight = "Low Birth Weight", PM25 = "PM2.5") %>% 
  select(CensusTract, TotalPopulation, CaliforniaCounty, LowBirthWeight, PM25, Poverty) %>% 
  mutate(LowBirthWeight = as.numeric(LowBirthWeight))
```

# Question a:
What is the average concentration of PM2.5 across all census tracts in California?
```{r}
mean_pm25 <- mean(data$PM25)
```

\noindent Answer: The average concentration of PM2.5 across all census tracts in California is `r round(mean_pm25, 2)` $\mu g/m^3$.

# Question b:
What county has the highest level of poverty in California?
```{r}
county_pov <- data %>%
  mutate(pov_per_capita = Poverty/TotalPopulation) %>% 
  group_by(CaliforniaCounty) %>% 
  summarise(mean_pov = mean(Poverty, na.rm = TRUE),
            mean_pov_per_capita = mean(pov_per_capita, na.rm = TRUE)) 

county_max_pov <- county_pov$CaliforniaCounty[which.max(county_pov$mean_pov)]
county_max_pov_per_capita <- county_pov$CaliforniaCounty[which.max(county_pov$mean_pov_per_capita)]
```

\noindent Answer: `r county_max_pov` County has the highest level of poverty in California based on mean poverty for all census tracts in each county. I decided not to use the county with the highest level of poverty per capital (which would have been `r county_max_pov_per_capita`) because according to the CES4 data dictionary the `Poverty` variable represents the percent of population living below two times the federal poverty level.

\noindent Note: there are 75 census tracts with NA values for poverty

# Question c:
Make a histogram depicting the distribution of percent low birth weight and PM2.5

```{r eval=TRUE, echo=FALSE, fig.height=5, fig.width=7}
low_birth_weight_histogram <- ggplot(data = data, aes(x = LowBirthWeight)) +
  geom_histogram(fill = "seagreen") +
  labs(x = "Percent Low Birth Rate")
```

<center>**Figure 1: Percent Low Birth Weight for California Census Tracts**</center>
```{r, eval=TRUE, echo=FALSE}
low_birth_weight_histogram
```

\noindent Figure 1 shows the distribution of percent low birth weights for California Census Tracts as reported by the California Office of Environmental Health Assessment (OEHHA). Data Source: CalEnviroScreen 4.0.

```{r , fig.width = 7, fig.height = 5, eval=TRUE, echo=FALSE}
pm25_histogram <- ggplot(data = data, aes(x = PM25)) +
  geom_histogram(fill = "steelblue3") +
  labs(x = "PM2.5 (micrograms per cubic meter)")
```

\newpage
<center>**Figure 2: Air Quality for California Census Tracts**</center>
```{r, eval=TRUE, echo=FALSE}
pm25_histogram
```

\noindent Figure 2 shows the distribution of annual mean PM 2.5 concentrations for California Census Tracts as reported by the California Office of Environmental Health Assessment (OEHHA). Data Source: CalEnviroScreen 4.0.

# Question d:
Estimate a OLS regression of `LowBirthWeight` on `PM25`. Report the estimated slope coefficient and its heteroskedasticity-robust standard error. Interpret the estimated slope coefficient. Is the effect of PM2.5 on low birth weight statistically significant at the 5% level?

$$\text{percent low birth weight}_i=\beta_{0}+\beta_{1} \cdot PM2.5 + \varepsilon_i$$
```{r}
ols_model_robust <- lm_robust(formula = LowBirthWeight ~ PM25, data = data)
```

```{r}
# use lm to estimate coefficients
ols_model <- lm(formula = LowBirthWeight ~ PM25, data = data)

# adjust standard errors using estimatr::starprep() instead of estimatr::lm_robust()
se_ols_model <- starprep(ols_model)
```


```{r, results = 'asis', echo = FALSE}
stargazer(ols_model, se = se_ols_model,
          type = "latex", ci=FALSE, no.space = TRUE, 
          header = FALSE, omit = c("Constant"), omit.stat = c("adj.rsq","ser", "f"),
          covariate.labels = c("PM 2.5 concentration"), dep.var.labels = c("percent low birth weight"),
          dep.var.caption = c(""), notes = c("Robust standard errors in parentheses"),
          title = "Low Birth Weight and PM 2.5 concentration", table.placement = "H")
```


```{r}
intercept <- ols_model_robust$coefficients[1]
intercept
```


```{r}
slope_coef <- ols_model_robust$coefficients[2]
slope_coef
```

```{r}
std_err <- ols_model_robust$std.error[2]
std_err
```
```{r}
ci_lower <- ols_model_robust$conf.low[2]
ci_lower

ci_upper <- ols_model_robust$conf.high[2]
ci_upper
```

\noindent Answer: The estimated slope coefficient for this model is `r round(slope_coef, 3)`. The estimated slope coefficient means that for a one unit increase in the concentration of PM2.5, percent low birth weight will increase by `r round(slope_coef, 3)`. The heteroskedasticity-robust standard error for this estimated slope coefficient is `r round(std_err, 3)`. Since the p-value is < 0.05, the effect of PM2.5 on percent low birth weight is statistically significant at the 5 % level.


```{r}
ols_model_plot <- ggplot(data = data, aes(x = PM25, y = LowBirthWeight)) +
  geom_point() +
  geom_smooth(method = lm)
```
\newpage
<center>**Figure 3: OLS Regression of Low Birth Weight on PM 2.5**</center>
```{r, eval=TRUE, echo=FALSE}
ols_model_plot
```

\noindent Figure 3 shows a positive correlation between low birth weight and PM 2.5 based on an Ordinary Least Squares Regression model of low birth weight on PM 2.5.

# Question f:
Add the variable Poverty as an explanatory variable to the regression in (d). Interpret the estimated coefficient on Poverty. What happens to the estimated coefficient on PM2.5, compared to the regression in (d). Explain.
```{r}
pm25_poverty_model_robust <- lm_robust(formula = LowBirthWeight ~ PM25 + Poverty, data = data)
```
```{r}
# use lm to estimate coefficients
pm25_poverty_model <- lm(formula = LowBirthWeight ~ PM25 + Poverty, data = data)

# adjust standard errors using estimatr::starprep() instead of estimatr::lm_robust
se_pm25_poverty_model <- starprep(pm25_poverty_model)
```

```{r , results = 'asis', echo = FALSE}
stargazer(ols_model, pm25_poverty_model, se = c(se_ols_model, se_pm25_poverty_model),
          type = "latex", ci=FALSE, no.space = TRUE, 
          header = FALSE, omit = c("Constant"), omit.stat = c("adj.rsq","ser", "f"),
          covariate.labels = c("PM 2.5 concentration", "Poverty"), dep.var.labels = c("percent low birth weight"),
          dep.var.caption = c(""), notes = c("Robust standard errors in parentheses"),
          title = "Low Birth Weight on PM 2.5 concentration and Poverty", table.placement = "H")
```

```{r}
poverty_coef <- pm25_poverty_model_robust$coefficients[3]
poverty_coef
```

```{r}
pm25_coef_2 <- pm25_poverty_model_robust$coefficients[2]
pm25_coef_2
```

\noindent The estimated coefficient on poverty is `r round(poverty_coef, 3)`. This mean that for a one unit increase in poverty, percent low birth weight will increase by `r round(poverty_coef, 3)` while holding PM2.5 constant. In this model, the estimated coefficient on PM2.5 is lower than the coefficient estimated with the model that did not include poverty. This suggests that PM2.5 is not the only variable to influences low birth weight.

# Question g:
From the regression in (f), test the null hypothesis that the effect of PM2.5 is equal to the effect of Poverty.

**null hypothesis:**
There is no difference in the estimated slope coefficients for PM2.5 and poverty.  

$$H_{0}: \beta_{1,PM2.5} - \beta_{2,poverty} = 0$$

**alternative hypothesis:** 
There is a difference in the estimated slope coefficient for PM2.5 and poverty.

$$H_{A}: \beta_{1,PM2.5} - \beta_{2,poverty} \neq 0$$

```{r}
linearHypothesis(model = pm25_poverty_model, hypothesis.matrix = c("PM25=Poverty"), white.adjust = "hc2")
```
\noindent Answer: Based on these results, we reject the null hypothesis that $\beta_{1,PM2.5} = \beta_{2,poverty}$.